\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amssymb,graphicx,booktabs,geometry}
\geometry{margin=1in}

\title{Modeling Locust Swarming with a Continuous-Activity ODE Model}
\author{Bar Jacobi \& Amy Vogel}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}

Locusts are well-known for their collective motion, where individuals alternate between solitary and gregarious phases. Modeling this behavior has been an active area of research, ranging from simple self-propelled particle (SPP) models to biologically detailed pause-and-go dynamics \cite{ariel2015locust}.  

In this project, we implement a minimal ODE-based model of locust collective motion. Each agent is represented by its position, heading, and an internal activity variable. We study how local neighbor interactions and stochasticity lead to the emergence of synchronized swarming. Our goals are to (i) situate the model in relation to existing work, (ii) clarify its assumptions and simplifications, (iii) define all parameters and variables with units, and (iv) evaluate robustness via parameter sweeps and heatmaps.

\section{Relation to Existing Models}

Ariel and Ayali (2015) review several frameworks for modeling locust collective motion:
\begin{itemize}
    \item The \textbf{Czirók model}, a 1D variant of the Vicsek self-propelled particle model, where individuals align to neighbors within a radius plus noise.
    \item The \textbf{Buhl model}, a refinement that adds persistence in direction, fitted to ring-arena experiments.
    \item The \textbf{Pause-and-go model}, where locusts switch between discrete moving and paused states, based on experimental observations of intermittent locomotion.
    \item \textbf{Escape-and-pursuit models}, motivated by cannibalism hypotheses, where individuals pursue those ahead and flee from those behind.
\end{itemize}

Our approach is closest to the Vicsek/Czirók family of models. The main distinction is that we introduce a continuous-valued activity variable $a \in [0,1]$ instead of a discrete pause/go state. This simplification allows us to explore threshold and saturation effects via a sigmoid activation function, while keeping the model tractable for robustness analysis.

\section{Model Definition}

Each locust is described by:
\begin{itemize}
    \item \textbf{Position:} $(x,y)$, coordinates in a periodic square domain of side length $L$ (m).
    \item \textbf{Heading:} $\theta$, orientation of movement (radians).
    \item \textbf{Activity:} $a \in [0,1]$, continuous locomotion variable (0 = paused, 1 = fully moving).
\end{itemize}

\subsection{Activity dynamics}
Activity relaxes toward a sigmoid function of neighbor count $n_i$:
\[
\frac{da_i}{dt} = -\frac{1}{\tau_a}\Big(a_i - \sigma(\beta_0 + \beta_1 n_i)\Big),
\]
where $\tau_a$ (s) is the relaxation timescale, $\beta_0$ is the baseline bias, and $\beta_1$ is the social activation gain. The sigmoid ensures thresholded, saturating influence of neighbors.

\subsection{Velocity}
Speed interpolates between minimum and maximum values depending on activity:
\[
v_i = v_{\min} + (v_{\max} - v_{\min}) a_i.
\]

\subsection{Heading dynamics}
Headings evolve by alignment with neighbors plus angular noise:
\[
\frac{d\theta_i}{dt} = \kappa \cdot \langle \sin(\theta_j - \theta_i)\rangle + \sqrt{2\eta}\,\xi,
\]
where $\kappa$ is alignment strength, $\eta$ is noise intensity (rad$^2$/s), and $\xi$ is Gaussian white noise.

\subsection{Domain}
Motion occurs in a square domain with periodic boundary conditions.

\section{Parameters and Variables}

\subsection{Mathematical parameters}

\begin{tabular}{@{}llll@{}}
\toprule
Symbol & Meaning & Units & Default \\
\midrule
$N$ & Number of locusts & count & 120 \\
$L$ & Domain size & m & 10 \\
$v_{\min}$ & Minimum speed & m/s & 0.01 \\
$v_{\max}$ & Maximum speed & m/s & 0.15 \\
$\kappa$ & Alignment strength & 1/s & 2.0 \\
$R$ & Vision radius & m & 1.0 \\
$\eta$ & Noise intensity & rad$^2$/s & 0.4 \\
$\beta_0$ & Baseline activity bias & --- & -1.0 \\
$\beta_1$ & Social activation gain & --- & 0.25 \\
$\tau_a$ & Activity relaxation time & s & 0.5 \\
\bottomrule
\end{tabular}

\subsection{Computational parameters}

\begin{tabular}{@{}llll@{}}
\toprule
Symbol & Meaning & Units & Default \\
\midrule
$dt$ & Simulation time step & s & 0.05 \\
$T$ & Total simulation time & s & 100 \\
seed & Random seed & --- & 42 \\
\bottomrule
\end{tabular}

\section{Robustness Analysis}

To test robustness, we performed 2D parameter scans. For each combination of parameters, we measured the \textbf{polarization}, defined as
\[
P = \sqrt{\Big(\frac{1}{N}\sum_i \cos \theta_i\Big)^2 + \Big(\frac{1}{N}\sum_i \sin \theta_i\Big)^2},
\]
which quantifies directional synchrony ($P=0$ = disorder, $P=1$ = perfect alignment).

\subsection{Scan 1: $\kappa$ vs $R$}
Polarization increases sharply when both alignment strength and vision radius exceed threshold values, reproducing the phase-transition–like behavior described in Vicsek-style models.

\subsection{Scan 2: $\kappa$ vs $\beta_1$}
Social activation enhances collective order: for high $\beta_1$, polarization rises at lower $\kappa$. This shows that intermittent movement dynamics, even when simplified to a continuous activity variable, strongly influence swarm coherence.

Heatmaps illustrating these results are included as figures (not shown in this text draft).

\section{Discussion}

Our model differs from prior work in that it uses a continuous activity variable rather than a binary pause/go state. This choice simplifies the dynamics while preserving threshold and saturation effects observed in experiments. Robustness analysis confirms that swarming arises only when alignment strength and social activation are sufficiently large, consistent with prior studies.

While our model does not capture all biological details (e.g., discrete pausing, escape-and-pursuit asymmetries), it highlights the minimal ingredients necessary for collective order. According to Ariel and Ayali (2015), such simplified models are valuable for understanding general principles of swarm dynamics, even if they abstract away some biological realism.

\section{Conclusion}

We implemented and analyzed a continuous-activity ODE model of locust swarming. Our work shows that simple local rules—alignment, social activation, and noise—are sufficient to generate robust collective motion. By situating this model within the broader literature, we demonstrate both how it simplifies existing frameworks and what insights it contributes. The approach also emphasizes the importance of robustness checks and parameter documentation, ensuring reproducibility and interpretability.

\begin{thebibliography}{9}
\bibitem{ariel2015locust}
Ariel, G., \& Ayali, A. (2015). Locust collective motion and its modeling. \emph{PLoS Computational Biology}, 11(12), e1004522.
\end{thebibliography}

\end{document}
